---
title: "432 Class 02"
author: "https://thomaselove.github.io/432-2024/"
date: "2024-01-18"
format: docx
---

## Today's Agenda

1.  Comparing Means
2.  Comparing Rates
3.  Fitting Linear Models
4.  Setting Up Lab 1, due Tuesday 2024-01-23 at Noon.

Course Notes: most relevant material in Chapters 1-5.

## Today's R Setup

```{r}
#| echo: true
knitr::opts_chunk$set(comment = NA)

library(broom) # for tidy, glance and augment
library(car) # for boxCox and vif
library(Epi) # for twoby2
library(GGally) # for ggpairs
library(MKinfer) # for boot.t.test
library(mosaic) # for favstats
library(naniar) # deal with missing values
library(nhanesA) # source of data
library(vcd) # for mosaic (plot) and assoc (plot)
library(janitor) # for tabyl and other things
library(tidyverse) # for all kinds of things

theme_set(theme_bw())
```

# Building an NHANES Data Set (see Course Notes Chapters 1-2)

## 2017 - March 2020 NHANES Data {.smaller}

1982 NHANES subjects ages 26-42 with complete data on these 9 variables:

Variable | Source | Description
-------- | ------ | ----------------------------------
SEQN | P-DEMO | Subject ID: Link (also in BPXO and HUQ)
age | P_DEMO | RIDAGEYR (restricted to ages 26-42 here)
educ | P_DEMO | DMDEDUC2 (five-category factor)
sbp1 | BPXO | BPXOSY1 = 1st Systolic BP reading, in mm Hg
sbp2 | BPXO | BPXOSY2 = 2nd Systolic BP reading
sbp3 | BPXO | BPXOSY3 = 3rd Systolic BP reading
sroh | HUQ | HUQ010 = five-categories E, VG, G, F, P
hospital | HUQ | HUQ071 = Yes or No
mentalh | HUQ | HUQ090 = Yes or No

## Building the Data (using nhanesA)

```{r}
#| echo: true
p_demo <- nhanes('P_DEMO') |>
  select(SEQN, RIDAGEYR, DMDEDUC2)

p_bpxo <- nhanes('P_BPXO') |>
  select(SEQN, BPXOSY1, BPXOSY2, BPXOSY3)

p_huq <- nhanes('P_HUQ') |>
  select(SEQN, HUQ010, HUQ071, HUQ090)

df_list <- list(p_demo, p_bpxo, p_huq)

nh_raw <- df_list |>
  reduce(left_join, by = 'SEQN') |>
  drop_na() |>
  filter(RIDAGEYR >= 26 & RIDAGEYR <= 42) |>
  as_tibble()
```

## Renaming and Cleaning Variables

```{r}
#| echo: true
nh1982 <- nh_raw |>
  rename(age = RIDAGEYR, educ = DMDEDUC2, 
         sbp1 = BPXOSY1, sbp2 = BPXOSY2, 
         sbp3 = BPXOSY3, sroh = HUQ010, 
         hospital = HUQ071, mentalh = HUQ090) |>
  replace_with_na_at(
    .vars = c("educ", "sroh", "hospital", "mentalh"), 
    condition = ~ .x %in% c("Refused", "Don't know")) |>
  mutate(educ = fct_recode(educ, 
             "Less than 9th Grade" = "Less than 9th grade",
             "9th - 11th Grade" = "9-11th grade (Includes 12th grade with no diploma)",
             "High School Grad" = "High school graduate/GED or equivalent",
             "Some College/AA" = "Some college or AA degree",
             "College Grad" = "College graduate or above"),
         sroh = fct_recode(sroh,
                "Excellent" = "Excellent,",
                "Very Good" = "Very good,",
                "Good" = "Good,",
                "Fair" = "Fair, or",
                "Poor" = "Poor?")) |>
  mutate(mean_sbp = (sbp1 + sbp2 + sbp3)/3,
         SEQN = as.character(SEQN)) |>
  drop_na() |> 
  droplevels()

write_rds(nh1982, "c02/data/nh1982.Rds")
```

## `nh1982`

```{r}
#| echo: true
glimpse(nh1982)
```

## Codebook (excerpt, without SEQN)

| Variable | Description (n = 1982)                                                                              |
|------------:|:----------------------------------------------------------|
|      age | Age in years (range 26-42, mean = 34)    |
|  meansbp | Mean SBP in mm Hg (range: 76 to 209, mean 116) |
| hospital | Yes if hospitalized in last 12m, else No (8% Yes) |
|  mentalh | Yes if saw a mental health professional in last 12m, else No (12% Yes) |
|     sroh | Self-reported Overall Health (see next slide) |
|     educ | Educational Attainment (see next slide) |

## SROH and Educational Attainment

```{r}
#| echo: true
nh1982 |> tabyl(sroh) |> adorn_pct_formatting()
nh1982 |> tabyl(educ) |> adorn_pct_formatting()

```

## Ingesting the Data (from .Rds)

If you don't want to work through the nhanesA import and tidying, you can simply work with the nh1982.Rds file provided on our 432-data page.

```{r}
#| echo: true
#| eval: false
nh1982 <- read_rds("c02/data/nh1982.Rds")

## not run here...
```

# Comparing Means (see Course Notes Chapter 3)

## Paired or Independent Samples?

In Analysis 1, we will compare the means of SBP1 and SBP2 for our 1982 participants.

In Analysis 2, we will compare the mean of SBP3 between our 159 participants who were hospitalized and the 1823 who were not?

Which of these analyses uses paired samples, and why?

## Paired Samples Analysis

```{r}
#| echo: true

nh1982 <- nh1982 |> mutate(SBP_diff = sbp1 - sbp2)

favstats(~ SBP_diff, data = nh1982)
```

```{r}
#| echo: true
#| eval: false

ggplot(nh1982, aes(sample = SBP_diff)) +
  geom_qq() + geom_qq_line(col = "red") +
  labs(title = "Normal Q-Q plot",
       subtitle = "SBP1 - SBP2 Differences",
       x = "Standard Normal Distribution",
       y = "Observed SBP1-SBP2 values")
```

## Paired SBP Differences

```{r}
#| echo: false
#| eval: true
#| fig-asp: 1

nh1982 <- nh1982 |> mutate(SBP_diff = sbp1 - sbp2)

ggplot(nh1982, aes(sample = SBP_diff)) +
  geom_qq() + geom_qq_line(col = "red") +
  labs(title = "Normal Q-Q plot",
       subtitle = "SBP1 - SBP2 Differences",
       x = "Standard Normal Distribution",
       y = "Observed SBP1-SBP2 values")
```

## Comparing Paired Samples

Want a 90% confidence interval for the true mean of the paired SBP1 - SBP2 differences:

-   t-based approach (same as linear model) assumes Normality
-   Wilcoxon signed rank approach doesn't assume Normality but makes inferences about the pseudo-median
-   bootstrap doesn't assume Normality, and describes mean

```{r}
#| echo: true
#| eval: false

set.seed(20240118)
boot.t.test(nh1982$SBP_diff, conf.level = 0.9, 
            boot = TRUE, R = 999)
```

Results on the next slide...

## Bootstrap for Mean of SBP1-SBP2 Differences

```{r}
#| echo: false
#| eval: true

set.seed(20240118)
boot.t.test(nh1982$SBP_diff, conf.level = 0.9, 
            boot = TRUE, R = 999)
```

## Comparing sbp3 by hospital: Independent Samples

```{r}
#| echo: true
favstats(sbp3 ~ hospital, data = nh1982) |>
  select(-missing) 
```

```{r}
#| echo: true
#| eval: false

ggplot(nh1982, aes(x = factor(hospital), y = sbp3)) +
  geom_violin(aes(fill = factor(hospital))) +
  geom_boxplot(width = 0.3, notch = TRUE) +
  guides(fill = "none") +
  labs(title = "SBP (3rd reading) by Hospitalization")
```

## SBP (3rd reading) vs. Hospitalization

```{r}
#| echo: false
#| eval: true
#| fig-align: center

ggplot(nh1982, aes(x = factor(hospital), y = sbp3)) +
  geom_violin(aes(fill = factor(hospital))) +
  geom_boxplot(width = 0.3, notch = TRUE) +
  guides(fill = "none") +
  labs(title = "SBP (3rd reading) by Hospitalization")
```

## Two Independent Samples, Comparing Means

Want a 90% confidence interval for the difference in means of SBP3 for people who were hospitalized - those who were not.

-   Pooled t-based approach (equivalent to linear model) assumes Normality and equal population variances
-   Welch t-based approach assumes Normality only
-   bootstrap assumes neither
-   Wilcoxon-Mann-Whitney rank sum assumes neither, but assesses a difference in locations, not the mean

## Pooled t test via linear model

```{r}
#| echo: true

lm2 <- lm(sbp3 ~ hospital, data = nh1982)

tidy(lm2, conf.int = TRUE, conf.level = 0.90) 

glance(lm2) |> select(r.squared, sigma) 
```

# Comparing Rates (see Course Notes, Chapter 4)

## A Two-by-Two Contingency Table

```{r}
#| echo: true

nh1982 |> tabyl(mentalh, hospital) |> 
  adorn_totals(where = c("row", "col")) |>
  adorn_title()
```

## Standard Epidemiological Format

```{r}
#| echo: true

nh1982 <- nh1982 |> 
  mutate(mentalh_f = fct_recode(factor(mentalh), 
                "Saw MHP" = "Yes", "No MHP" = "No"),
         mentalh_f = fct_relevel(mentalh_f, 
                "Saw MHP", "No MHP"),
         hospital_f = fct_recode(factor(hospital), 
                "Hosp." = "Yes", "No Hosp." = "No"),
         hospital_f = fct_relevel(hospital_f, 
                "Hosp.", "No Hosp."))

nh1982 |> tabyl(mentalh_f, hospital_f)
```

## Two by Two Table Analysis

```{r}
#| echo: true
twoby2(nh1982$mentalh_f, nh1982$hospital_f, conf.level = 0.90)
```

## A Larger Two-Way Table

What is the association of Educational Attainment with Self-Reported Overall Health?

```{r}
#| echo: true

nh1982 |> tabyl(educ, sroh) |> 
  adorn_totals(where =c("row","col"))|> adorn_title()
```

## Our 5x5 Table, showing SROH Proportions

```{r}
#| echo: true
nh1982 |> tabyl(educ, sroh) |> 
  adorn_totals(where = c("row")) |>
  adorn_percentages(denominator = "row") |> 
  adorn_pct_formatting() |> adorn_title()
```

## Mosaic Plot for our 5x5 Table

```{r}
#| echo: true
#| fig-height: 5
mosaic(~ educ + sroh, data = nh1982, highlighting = "sroh")
```

## Pearson $\chi^2$ test for our 5x5 Table

```{r}
#| echo: true

chisq.test(xtabs(~ educ + sroh, data = nh1982))
```

## Association Plot for our 5x5 Table

```{r}
#| echo: true
#| fig-height: 5
assoc(~ educ + sroh, data = nh1982)
```

# Fitting Linear Models (see Course Notes, Chapter 5)

## We'll fit two models today

1.  Predict mean SBP using Age alone.
2.  Predict mean SBP (across three readings) using Age, Self-Reported Overall Health Status and Hospitalization Status.

```{r}
#| echo: true
temp_mod1 <- lm(mean_sbp ~ age, data = nh1982)
temp_mod2 <- lm(mean_sbp ~ age + sroh + hospital, 
                data = nh1982)
```

Note that I'm not doing any predictive validation today (remember that I did that in Class 1), so I won't split the sample.

## Box-Cox Plot to suggest potential outcome transformations

```{r}
#| echo: true
#| fig-height: 4.5
boxCox(temp_mod2)

nh1982 <- nh1982 |> mutate(inv_sbp = 1000/mean_sbp)
```

## Scatterplot Matrix (from `ggpairs()`)

```{r}
#| echo: true
#| fig-height: 4.5
ggpairs(nh1982, columns = c(2, 7, 8, 14), switch = "both",
        lower=list(combo=wrap("facethist", bins=20)))
```

## Checking Collinearity: Variance Inflation Factors

```{r}
#| echo: true
vif(lm(inv_sbp ~ age + sroh + hospital, data = nh1982))
```

## Tidied Coefficients for Model `m1`

```{r}
#| echo: true

m1 <- lm(inv_sbp ~ age, data = nh1982)

tidy(m1, conf.int = TRUE, conf.level = 0.9)
```

## Tidied Coefficients for Model `m2`

```{r}
#| echo: true

m2 <- lm(inv_sbp ~ age + sroh + hospital, data = nh1982)

tidy(m2, conf.int = TRUE, conf.level = 0.9)
```

## Fit Summaries for Models `m1` and `m2`

```{r}
#| echo: true
bind_rows(glance(m1), glance(m2)) |>
  mutate(model = c("m1", "m2")) |> 
  select(model, r2 = r.squared, adjr2 = adj.r.squared, 
         sigma, AIC, BIC, nobs, df, df.residual) 
```

Which model appears to fit the data better?

## Residual Plots for Model `m2`

```{r}
par(mfrow = c(2,2)); plot(m2); par(mfrow = c(1,1))
```

## Making a Prediction in New Data

Suppose a new person is age 29, was not hospitalized, and their SROH is "Good". What is their predicted mean systolic blood pressure?

-   Our models predict 1000/mean_sbp and augment places that prediction into `.fitted`.
-   To invert, divide `.fitted` by 1000, then take the reciprocal of that result. That's just 1000/`.fitted`.

## Making a Prediction in New Data

```{r}
#| echo: true
new_person <- tibble(age = 29, sroh = "Good", hospital = "No")
bind_rows(augment(m1, newdata = new_person), 
          augment(m2, newdata = new_person)) |>
  mutate(model = c("m1", "m2"), fit_meansbp = 1000/.fitted) |>
  select(model, fit_meansbp, .fitted, age, sroh, hospital) 
```

# Setting Up Lab 1, due Tuesday 2024-01-23 at Noon

## Lab 1 Question 1

I provide some County Health Rankings data for Ohio's 88 counties. You create a visualization involving information from at least three different variables using R and Quarto.

-   Include proper labels and a meaningful title.
-   Include a caption (75 words or fewer) that highlights the key result.
-   What is the question you are trying to answer with this visualization?

There is a Quarto template for Lab 1, in addition to the data set.

## Lab 1 Question 2

Create a linear regression model to predict `obese_pct` as a function of `food_env` and `median_income` (all of these are quantitative variables.)

a.  Specify and fit the model, interpret `food_env` coefficient.
b.  Evaluate quality of model in terms of adherence to regression assumptions via four key residual plots.
c.  Build a nice table comparing your model to a simple regression for `obese_pct` using only `food_env`, and then reflect on your findings.

## Next Week?

-   Lab 1 due Tuesday Noon 
    - Answer Sketch available Wednesday
-   Developing Inferences Using Survey Weights
-   Linear Regression and ANOVA/ANCOVA models
