---
title: "Building Table 1 from the `bradley.csv` data"
author: "Thomas E. Love"
date: today
format: 
  html:
    toc: true
    number-sections: true
    code-fold: show
    code-tools: true
    code-overflow: wrap
    embed-resources: true
    date-format: iso
    theme: materia  ## change the theme if you prefer
---

This is a "live code" example that Dr. Love developed for Class 12 of 432, on 2023-02-23.

## Load Packages

```{r}
#| message = FALSE
knitr::opts_chunk$set(comment = NA)

library(tableone)
library(janitor)
library(tidyverse)
```

## Ingest the `bradley.csv` data 

```{r}
brad_raw <- read_csv("data/bradley.csv", 
                     show_col_types = FALSE) |>
    clean_names()

brad_raw
```

## Attempt 1

```{r}
vars <- c("age", "sex", "race_eth", "married", "location")
trt <- c("status")

attempt_1 <- CreateTableOne(data = brad_raw, 
                       vars = vars, 
                       strata = trt)
print(attempt_1)
```

## What's wrong here?

```{r}
brad_new <- brad_raw |>
    mutate(marital = fct_recode(factor(married), 
                                "yes" = "1", 
                                "no"  = "0")) |>
    mutate(loc = fct_relevel(location, 
                             "ICU", "Bed", "Other"))
```

## Second Attempt

```{r}
vars <- c("age", "sex", "race_eth", "marital", "loc")
factorvars <- c("sex", "race_eth", "marital", "loc")
trt <- c("status")

attempt_2 <- CreateTableOne(data = brad_new, 
                       vars = vars, 
                       factorVars = factorvars,
                       strata = trt)
print(attempt_2)
```

## Show alternative summaries?

```{r}
print(attempt_2, 
      nonnormal = c("age"),
      exact = c("sex", "race_eth", "marital"))
```

## A more detailed summary?

```{r}
summary(attempt_2)
```

## Send original attempt 2 to Excel for polish?

One option is to **save the Table 1** to a `.csv` file, which you can then open directly in Excel. This is the approach I generally use. Note the addition of some `quote`, `noSpaces` and `printToggle` selections here.

```{r}
brad_t1 <- print(attempt_2, 
      quote = FALSE, noSpaces = TRUE, printToggle = FALSE)

write.csv(brad_t1, 
          file = "data/bradley_table1_result.csv")
```

You can then open the `bradley_table1_result.csv` file in Excel and edit further.

